Niche complementarity between pollinators increases community-level plant reproductive success

### 2019-08-08 Network_structure_and_function.Rmd
rmd

  Here, we present one of the first efforts linking pollinator visitation and plant reproductive success at the community level using empirical data on plant-pollinator interaction networks and plant reproductive success. To this end, we use a well-replicated dataset encompassing 16 well-resolved plant-pollinator interaction networks coupled with data on the reproductive success of 19 plant species recorded in Mediterranean shrub ecosystems. Our study focuses on understanding whether adding information on selected interaction network structure indices to previously used simple visitation metrics (e.g., the number and diversity of pollinator species visiting a plant species) aids in better explaining the differences observed in community-wide reproductive success. In doing so, we conducted our analyses focusing on reproductive success at two different levels: (i) at the species level by considering the effect of the position of a focal species within the larger network and its impact on its individual reproductive success, and (ii) at the site level, by evaluating how attributes that describe the whole site might affect average values of reproductive success. Specifically, our study focuses on how the interplay between niche complementarity and redundancy determines reproductive success. Plant reproductive success requires of the delivery of conspecific pollen and thus of a certain degree of niche complementarity [@Bluthgen2011]. Yet, greater values of nestedness, which imply redundancy in species functions, are thought to promote species diversity [@Bastolla2009] and stability [@Thebault2010a] within plant-polllinator networks. At present, we do not know how either of these network characteristics affects the functions performed by pollinators. Finally, in addition to average values, we also evaluate whether network structure helps explain differences in equity in reproductive success across species within a community, as a measure of evenness in the pollination service delivered.

@@

#calculate sorensen index to look at beta-diversity across sites for plants
library(betapart)
d.sub<-subset(d, d$Out=="transect") #keep only plants surveyed in transects, remove focals
d.beta<-table(d.sub$Site_ID, d.sub$Plant_gen_sp)
d.beta[d.beta>1]<-1 #make it binary
d.betadiv<-beta.multi(d.beta)
sor<-d.betadiv[3]

#Alternativelly, as I am not understanding betapart.
#using vegan classic apporach.
mean(betadiver(d.beta, method = 11)) #this is theoretically the mean of soerensens across sites according to vegan. A bit lower. 0.41.
summary(betadiver(d.beta, method = 11)) #max value 0.66.
```

@@

Our study was conducted in SW Spain within the area of influence of Doñana National Park, i.e., within the limits of the Natural Space of Doñana as defined by the local government (Junta de Andalucía, Fig. 1). All sites were located within similar elevations (ranging from 50 to 150 m a.s.l.), similar habitat and soil types, and presented similar plant compositions (plant mean Sørensen beta-diversity among sites = `r round(sor$beta.SOR, 2)`), reducing potential confounding factors. Here, we surveyed 16 Mediterranean woodland patches with an average distance of 7 km between them (min= 3 km, max= 46.5 km). Each site was surveyed 7 times during the flowering season of 2015 (from February to May) following a 100-m x 2 m transect for 30 mins. Along each transect, we identified all plant species and recorded all the floral visitors that landed on their flowers during each 30-min period. Only floral visitors (from now on referred to as pollinators) that could not be identified in the field were captured, stored and identified in the laboratory by FPM with the aid of experts in the different taxonomic groups (see acknowledgements). In addition, at each round we conducted 3 minutes of focal observations recording all floral visitors observed on 3 plant individuals per species belonging to the 19 most common (based on previous surveys) plant species across the study area (mean $\pm$ SD: `r round(mean(foc3$tot.species), digits=2)` $\pm$ `r round(sd(foc3$tot.species), digits=2)` species per site). Furthermore, we included some interactions between plant and pollinator individuals that were not observed during the sampling but that were opportunistically recorded immediately before or after the sampling periods, as some of these interactions are difficult to document and might be important to define network structure [@Jordano2016]. These opportunistic interactions represented `r round(table(d$Out)[2]/(table(d$Out)[1]+table(d$Out)[2]+table(d$Out)[3])*100, digits=2)` of all interactions recorded. All surveys were done under similar weather conditions, avoiding windy or rainy days. Surveys were done during mornings and afternoons with the sampling order being established randomly.

@@

Within each site, we marked between 3 and `r max(f.inds$tot.inds)` individuals (mean $\pm$ SD: `r round(mean(f.inds$tot.inds), digits=2)` $\pm$ `r round(sd(f.inds$tot.inds), digits=2)`, Table S2) belonging to `r min(tableS1$tot.sps)` to `r max(tableS1$tot.sps)` plant species (mean $\pm$ SD:`r round(mean(tableS1$tot.sps), digits=2)` $\pm$ `r round(sd(tableS1$tot.sps), digits=2)`, Table S1), depending on the availability and presence of flowers during the sampling events. For each individual, at the end of the season, we 
recorded fruit set (i.e. the proportion of flowers that set fruit), the average number of seeds per fruit and the average fruit and seed weight per fruit (`r min(s.fruitnum$tot.fruits)`-`r max(s.fruitnum$tot.fruits)` fruits subsampled, mean $\pm$ SD: `r round(mean(s.fruitnum$tot.fruits), digits=2)` $\pm$ `r round(sd(s.fruitnum$tot.fruits), digits=2)`, Table S3). Our survey included a total of `r nlevels(f2$Plant.sp)` different plant species across our `r nlevels(f2$Site_ID)` sites. Plants species were selected based on their availability, with sampling being focused on the most abundant plant species. The values at the species level were then averaged per site to calculate unique reproductive success measures at the site level. 68% of the plant species depend on pollinators to maximize their reproduction (Table S4).

@@

  In order to analyse how differences in network structure might affect plant reproductive success, we constructed plant-pollinator interaction networks by pooling the data for the 7 rounds of sampling. We thus obtained one interaction network per site, representing the incidence of visits by different pollinator species to different plant species. For each network, we then proceeded to extract a series of relevant network metrics at the species and site levels.

  In addition, we checked for potential spatial autocorrelation in our data by means of Mantel correlograms. Autocorrelation values were low for all variables included in our analyses (Figure S1).

@@

  At the species level, response variables included the fruit set for different individuals of each species analyzed using a binomial distribution, the average number of seeds per fruit analyzed using a normal distribution, and the average values of fruit and seed weight fitted to Poisson distributions. The number of seeds per fruit was centered and scaled (i.e., we subtracted column means and divided by standard deviation) to allow meaningful comparisons across species with contrasting life histories. As explanatory variables, model 1 included pollinator richness, and the total number of visits received by each plant species; while model 2 added the two network attributes calculated at the species level: average plant niche overlap and centrality. For both models, we included plant species nested within site and site as random effects to account for the non-indepedence of several individuals measured per species and site.

	At the site level, we upscaled our species-level analyses. As response variables we had the average reproductive success per site (i.e., average fruit set analyzed using a binomial distribution, average number of seeds per fruit and average fruit and seed weight using a normal distribution). We thus had a single value per site and no random effects are needed in this case. In this case, model 1 included total pollinator richness and total pollinator abundance (i.e. number of visits received by all plants within the community) as explanatory variables. Model 2, in turn, added information on network structure by including nestedness and pollinator niche complementarity as explanatory variables.

@@

  To test this, we ran a simple simulation in which the number of species considered increased at each step and for each step we re-calculated equity in reproductive success. Instead of drawing plant species randomly for each step, we tested all possible combinations for each plant number level and network, as the number of combinations is small (e.g. for n = 3 plant selected out of 6 there is only 20 possible combinations). Then, we tested if the relationship between equity in reproductive success and functional complementarity (given its importance in determining differences in reproductive success, see Results section) changes as a function of the number of plants considered within our simulated communities. To this end, for each level of species number considered, we randomly selected one of the generated equity values across each of the 16 communities and regressed these 16 values against our network level predictor and extracted the model slope estimates. We repeated this process 1,000 times and averaged all slope estimates. We expect that the more plants considered, the larger the resulting average estimates will be. Note that we only interpret the mean effects, as the variance among different plant number of species considered depends on the initial number of combinations.

@@

At the species level, in the case of fruit set, our results show that model 2 shows the best fit to our data (lowest AIC value), explaining `r round(r2$Marginal[1], 2)*100`% of the variability observed. In this case, we find a positive effect of a network structure metric, the centrality of a focal plant within the overall network on its fruit set (Table 1, Fig. 2A).

@@

 For the average number of seeds per fruit, our results show again that model 2 shows the best fit, explaining `r round(r2b$Marginal[1], 2)*100`% of the variability observed in our data. In this case, we find a positive effect of the niche overlap between plant species on the number of seeds produced (Table 1B, Fig. 2B).

@@

When evaluating the effect of differences in community composition and network structure for equity in reproductive success across the different species within a community we find that model 1 is the best model for all the thresholds considered (50^th, 25^th and 75^th percentiles). Here, none of the variables considered are able to explain differences observed in equity across sites (Tables S9, S10, S11).

@@

  One of the first conclusions we can extract from the fact that in most cases both of the models we considered (i.e., the simple model based on visitation metrics and the more complex one including network structure metrics) were equally good, is that the added complexity of measuring the full network of interactions may not pay off for rapid assessments. Hence, simple visitation metrics, such as pollinator richness, might be enough [@Garibaldi2013; @Garibaldi2015]. Yet, adding network level information may inform us of the potential ecological mechanisms underlying the processes driving these observed patterns.

@@

  In the case of equity, we did not find a significant effect of either simple visitation or network structure metrics. However, the results of our simulation on the importance of network structure as the number of plant species considered increases, shows us that this effect increases when more than four plant species are considered. This implies that if we were able to measure reproductive success for all the plant species in all the communities (which is not feasible given constraints in sampling effort), we might find that the effects of network structure on equity might be more prevalent.

@@

  In summary, our findings show that the analysis of natural communities of interacting species using network analysis not only represents an ideal way of visualizing and grasping the complexity present within these communities. Rather, it also represents a manner of mechanistically understanding differences observed across the reproductive success of individuals and/or species while linking them to potential ecological mechanisms. Future studies could build on this body of research by focusing on the long-lasting effects of network structure for plant fitness, by evaluating the consequences of different network structures on other stages of a plant´slife cycle (e.g., seed viability) and whether these differences are passed on to later generations (e.g., though different survival rates).

@@

The authors would like to thank Oscar Aguado for help in identifying pollinator species. We would also like to acknowledge the efforts of the editor Cédric Gaucherel, as well as those of Michael Lattorff, Nicolas Deguines, and three anonymous reviewers who provided very helpful comments and suggestions in a previous version of this manuscript. AM received funding from a Juan de la Cierva (IJCI-2014-22558) and Ikerbasque fellowships. IB acknowledges funding from MSC-PCIG14-GA-2013-631653 BeeFun Project. We thank Doñana’s Singular Scientific-Technical Infrastructure (ICTS-RBD) for access to the park.

>>>>

  Here, we present one of the first efforts linking pollinator visitation and plant reproductive success at the community level using empirical data on plant-pollinator interaction networks and plant reproductive success. To this end, we use a well-replicated dataset encompassing 16 well-resolved plant-pollinator interaction networks coupled with data on the reproductive success of 19 plant species recorded in Mediterranean shrub ecosystems. Our study focuses on understanding whether adding information on selected interaction network structure indices to previously used simple visitation metrics (e.g., the number and diversity of pollinator species visiting a plant species) aids in better explaining the differences observed in community-wide reproductive success. In doing so, we conducted our analyses focusing on reproductive success at two different levels: (i) at the species level by considering the effect of the position of a focal species within the larger network and its impact on its individual reproductive success, and (ii) at the site level, by evaluating how attributes that describe the whole site might affect average values of reproductive success. Specifically, our study focuses on how the interplay between niche complementarity and redundancy determines reproductive success. Plant reproductive success requires of the delivery of conspecific pollen and thus of a certain degree of niche complementarity [@Bluthgen2011]. Yet, greater values of nestedness, which imply redundancy in species functions, are thought to promote species diversity [@Bastolla2009] and stability [@Thebault2010a] within plant-pollinator networks. At present, we do not know how either of these network characteristics affects the functions performed by pollinators. Finally, in addition to average values, we also evaluate whether network structure helps explain differences in equity in reproductive success across species within a community, as a measure of evenness in the pollination service delivered.

@@

#calculate sorensen index to look at beta-diversity across sites for plants
library(betapart)
d.sub<-subset(d, d$Out=="transect") #keep only plants surveyed in transects, remove focals
d.beta<-table(d.sub$Site_ID, d.sub$Plant_gen_sp)
d.beta[d.beta>1]<-1 #make it binary
d.betadiv<-mean(betadiver(d.beta, method = 11)) 
#summary(betadiver(d.beta, method = 11)) #max value 0.66.
```

@@

Our study was conducted in SW Spain within the area of influence of Doñana National Park, i.e., within the limits of the Natural Space of Doñana as defined by the local government (Junta de Andalucía, Fig. 1). All sites were located within similar elevations (ranging from 50 to 150 m a.s.l.), similar habitat and soil types, and presented similar plant composition (plant mean Sørensen beta-diversity among sites = `r round(d.betadiv, 2)`), reducing potential confounding factors. Here, we surveyed 16 Mediterranean woodland patches with an average distance of 7 km between them (min= 3 km, max= 46.5 km). Each site was surveyed 7 times during the flowering season of 2015 (from February to May) following a 100-m x 2 m transect for 30 mins. Along each transect, we identified all plant species and recorded all the floral visitors that landed on their flowers during each 30-min period. Only floral visitors (from now on referred to as pollinators) that could not be identified in the field were captured, stored and identified in the laboratory by FPM and experts in the different taxonomic groups (see acknowledgements). In addition, at each round we conducted 3 minutes of focal observations recording all floral visitors observed on 3 plant individuals per species belonging to the 19 most common (based on previous surveys) plant species across the study area (mean $\pm$ SD: `r round(mean(foc3$tot.species), digits=2)` $\pm$ `r round(sd(foc3$tot.species), digits=2)` species per site). Furthermore, we included some interactions between plant and pollinator individuals that were not observed during the sampling but that were opportunistically recorded immediately before or after the sampling periods, as some of these interactions are difficult to document and might be important to define network structure [@Jordano2016]. These opportunistic interactions represented `r round(table(d$Out)[2]/(table(d$Out)[1]+table(d$Out)[2]+table(d$Out)[3])*100, digits=2)` of all interactions recorded. All surveys were done under similar weather conditions, avoiding windy or rainy days. Surveys were done during mornings and afternoons with the sampling order being established randomly.

@@

Within each site, we marked between 3 and `r max(f.inds$tot.inds)` individuals (mean $\pm$ SD: `r round(mean(f.inds$tot.inds), digits=2)` $\pm$ `r round(sd(f.inds$tot.inds), digits=2)`, Table S2) belonging to `r min(tableS1$tot.sps)` to `r max(tableS1$tot.sps)` plant species (mean $\pm$ SD:`r round(mean(tableS1$tot.sps), digits=2)` $\pm$ `r round(sd(tableS1$tot.sps), digits=2)`, Table S1), depending on the availability and presence of flowers during the sampling events. For each individual, at the end of the season, we 
recorded fruit set (i.e. the proportion of flowers that set fruit), the average number of seeds per fruit and the average fruit and seed weight per fruit (`r min(s.fruitnum$tot.fruits)`-`r max(s.fruitnum$tot.fruits)` fruits subsampled, mean $\pm$ SD: `r round(mean(s.fruitnum$tot.fruits), digits=2)` $\pm$ `r round(sd(s.fruitnum$tot.fruits), digits=2)`, Table S3). Our survey included a total of `r nlevels(f2$Plant.sp)` different plant species across our `r nlevels(f2$Site_ID)` sites. Plants species were selected based on their availability, with sampling being focused on the most abundant plant species. The values at the species level were then averaged per site to calculate unique reproductive success measures at the site level. All plant species depend on pollinators to maximize their reproduction (Table S4).

@@

  In order to analyse how differences in network structure might affect plant reproductive success, we constructed plant-pollinator interaction networks by pooling the data for the 7 rounds of sampling. We thus obtained one interaction network per site, representing the number of individuals of different pollinator species recorded visiting each different plant species. For each network, we then proceeded to extract a series of relevant network metrics at the species and site levels.

  In addition, we checked for potential spatial autocorrelation in our data by means of Mantel correlograms. Autocorrelation values were low for all variables included in our analyses (Figure S1) and hence we treat each site as independent in our analysis.

@@

  At the species level, response variables included the fruit set for different individuals of each species analyzed using a binomial distribution, the average number of seeds per fruit analyzed using a normal distribution, and the average values of fruit and seed weight fitted to Poisson distributions. The number of seeds per fruit was centered and scaled (i.e., we subtracted column means and divided by standard deviation) to allow meaningful comparisons across species with contrasting life histories. As explanatory variables, model 1 included pollinator richness, and the total number of visits received by each plant species; while model 2 added the two network attributes calculated at the species level: average plant niche overlap and centrality. For both models, we included plant species nested within site and site as random effects to account for the non-independence of several individuals measured per species and site.

	At the site level, we upscaled our species-level analyses. As response variables we had the average reproductive success per site (i.e., average fruit set analyzed using a binomial distribution, average number of seeds per fruit and average fruit and seed weight using a normal distribution). We thus had a single value per site and no random effects are needed. In this case, model 1 included total pollinator richness and total pollinator abundance (i.e. number of visits received by all plants within the community) as explanatory variables. Model 2, in turn, added information on network structure by including nestedness and pollinator niche complementarity as explanatory variables.

@@

  To test this, we ran a simple simulation in which the number of species considered increased at each step and for each step we re-calculated equity in reproductive success. Instead of drawing plant species randomly for each step, we tested all possible combinations for each plant number level and network, as the number of combinations is small (e.g. for n = 3 plant selected out of 6 there is only 20 possible combinations). Then, we tested if the relationship between equity in reproductive success and functional complementarity (given its importance in determining differences in reproductive success, see Results section) changes as a function of the number of plants considered within our simulated communities. To this end, for each level of species number considered, we randomly selected one of the generated equity values across each of the 16 communities and regressed these 16 values against our network level predictor and extracted the model slope estimates. We repeated this process 1,000 times and averaged all slope estimates. We expect that the more plants considered, the larger the resulting average estimates will be. Note that we only interpret the mean effects, as the variance among different plant number of species considered depends on the initial number of possible combinations.

@@

At the species level, in the case of fruit set, our results show that model 2 shows the best fit to our data (lowest AIC value), and fixed  effects explains `r round(r2$Marginal[1], 2)*100`% of the variability observed. In this case, we find a positive effect of a network structure metric, the centrality of a focal plant within the overall network on its fruit set (Table 1, Fig. 2A).

@@

 For the average number of seeds per fruit, our results show again that model 2 shows the best fit, with fixed effects explaining `r round(r2b$Marginal[1], 2)*100`% of the variability observed in our data. In this case, we find a positive effect of the niche overlap between plant species on the number of seeds produced (Table 1B, Fig. 2B).

@@

When evaluating the effect of differences in community composition and network structure for equity in reproductive success across the different species within a community we find that model 1 is the best model for all the thresholds considered (50^th, 25^th and 75^th percentiles). However, none of the variables considered are able to explain differences observed in equity across sites (Tables S9, S10, S11).

@@

  One of the first conclusions we can extract from the fact that in most cases both of the models we considered (i.e., the simple model based on visitation metrics and the more complex one including network structure metrics) were equally good, is that the added complexity of measuring the full network of interactions may not pay off for rapid assessments. Hence, simple visitation metrics, such as pollinator richness, might be enough to describe general patterns [@Garibaldi2013; @Garibaldi2015]. Yet, adding network level information may inform us of the potential ecological mechanisms underlying the processes driving these observed patterns.

@@

  In the case of equity, we did not find a strong effect of either simple visitation or network structure metrics. However, the results of our simulation on the importance of network structure as the number of plant species considered increases, shows us that this effect increases when more than four plant species are considered. This implies that if we were able to measure reproductive success for all the plant species in all the communities (which is not feasible given constraints in sampling effort), we might find that the effects of network structure on equity might be more prevalent.

@@

  In summary, our findings show that the analysis of natural communities of interacting species using network analysis not only represents an ideal way of visualizing and grasping the complexity present within these communities. Rather, it also represents a manner of mechanistically understanding differences observed across the reproductive success of individuals and/or species while linking them to potential ecological mechanisms.

@@

The authors would like to thank Oscar Aguado for identifying pollinator species. We would also like to acknowledge the efforts of the editor Cédric Gaucherel, as well as those of Michael Lattorff, Nicolas Deguines, and three anonymous reviewers who provided very helpful comments and suggestions in a previous version of this manuscript. AM received funding from a Juan de la Cierva (IJCI-2014-22558) and Ikerbasque fellowships. IB acknowledges funding from MSC-PCIG14-GA-2013-631653 BeeFun Project. We thank Doñana’s Singular Scientific-Technical Infrastructure (ICTS-RBD) for access to the park.
